One page description

### 2016-12-07 publications/Journal/1-introduction.tex
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything
Rewrote first part of the paper to the new synchronization terminology

But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
Whereas \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, its inherent parallelism is often insufficient~\cite{Himmelspach}.
Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to (\textsf{Parallel}) \textsf{DEVS} simulation~\cite{globaltimewarp}.
With synchronization protocols, different simulation cores can be at different points in simulated time, significantly increasing parallelism, at the cost of synchronization overhead.
While several parallel \textsf{DEVS} simulation kernels exist, they are often limited to a single synchronization protocol.
The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other for specific models~\cite{Jafer}.
The parallel simulation capabilities of current tools are therefore limited to specific domains.

This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model: users only determine the protocol to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can run the exact same model using the exact same tool, but with different synchronization protocols.
As model behaviour, and thus the ideal synchronization protocol, might change during simulation, runtime switching of synchronization protocols is also supported.
This runtime switching can be based on performance metrics, which are logged during simulation.
Information is made available to a separate component, where a choice can be made about which synchronization protocol to use. % It's manual now, can be extended to full automatic.
Additionally, we investigate how model allocation influences the performance of our synchronization protocols.
To this end, we have included an allocation component in our simulation kernel.

@@

Using several benchmark models, we demonstrate the factors influencing the performance under a given synchronization protocol.
Additionally, we investigate a model which changes its behaviour (and ideal synchronization protocol) during simulation.
\textit{Dxex}, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
With \textit{dxex} users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own, or variants of existing ones.
To verify that this modularity does not hamper performance, we compare to \textit{adevs}~\cite{adevs}, another \textsf{Parallel DEVS} simulation tool.

>>>>

But with the ever increasing complexity of simulation models, parallel simulation becomes necessary to perform the simulation within reasonable time bounds.
Whereas \textsf{Parallel DEVS}~\cite{ParallelDEVS} was introduced to increase parallelism, its inherent parallelism is often insufficient~\cite{Himmelspach}.
Several synchronization protocols from the discrete event simulation community~\cite{FujimotoBook} have been applied to (\textsf{Parallel}) \textsf{DEVS} simulation~\cite{globaltimewarp}.
With synchronization protocols, different simulation kernels can be at different points in simulated time, significantly increasing parallelism at the cost of synchronization overhead.
While several parallel \textsf{DEVS} simulation tools exist, they are often limited to a single synchronization protocol.
The reason for different synchronization protocols, however, is that their distinct nature makes them applicable in different situations, each outperforming the other for specific models~\cite{Jafer}.
The parallel simulation capabilities of current tools are therefore limited to specific domains.

This paper introduces \textit{DEVS-Ex-Machina}\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``\textit{dxex}''): our simulation tool~\cite{dxex} which offers 6 synchronization protocols.
The simulation tool can spawn multiple kernels, each of which has its own simulation time and simulation control, though they may exchange timestamped events with each other.
Each kernel can spawn multiple threads, each of which will be mapped to a physical CPU core by the operating system.
We note the presence of two kinds of synchronization: inter-kernel and intra-kernel.
Inter-kernel synchronization uses either no synchronization, conservative synchronization, or optimistic synchronization.
Intra-kernel synchronization uses either no synchronization or exploits the inherent parallelism of Parallel DEVS to allocate concurrent transitions to separate threads.
The inherent parallelism in the intra-kernel synchronization protocol is used \pSim in the remainder of this paper.

This results in the following set of synchronization protocol combinations:
\begin{itemize}
\item No inter-kernel synchronization, no intra-kernel synchronization (\textit{NN})
\item Conservative inter-kernel synchronization, no intra-kernel synchronization (\textit{CN})
\item Optimistic inter-kernel synchronization, no intra-kernel synchronization (\textit{ON})
\item No inter-kernel synchronization, \pSim intra-kernel synchronization (\textit{NP})
\item Conservative inter-kernel synchronization, \pSim intra-kernel synchronization (\textit{CP})
\item Optimistic inter-kernel synchronization, \pSim intra-kernels synchronization (\textit{OP})
\end{itemize}

When refering to a set of them, we use the notation \textit{x} to denote any possible algorithm.
For example, \textit{xP} refers to NP, CP, and OP.

The selected synchronization protocol is transparent to the simulated model: users only determine the protocol to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can keep using the exact same tool, but with different synchronization protocols.
As model behaviour, and thus the ideal synchronization protocol, might change during simulation, runtime switching of synchronization protocols is also supported.
This runtime switching can be based on performance metrics, which are logged during simulation.
Information is made available to a separate component, where a choice can be made about which synchronization protocol to use. % It's manual now, can be extended to full automatic.
Additionally, we investigate how model allocation influences the performance of our synchronization protocols.
To this end, we have included an allocation component in our simulation kernel.

@@

Using several benchmark models, we demonstrate the factors influencing the performance under a given synchronization protocol.
Additionally, we investigate a model which changes its behaviour (and corresponding ideal synchronization protocol) during simulation.
\textit{Dxex}, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol.
With \textit{dxex} users can always opt to use the fastest protocol available, and through its modularity, users could even implement their own, or variants of existing ones.
To verify that this modularity does not hamper performance, we compare to \textit{adevs}~\cite{adevs}, another \textsf{Parallel DEVS} simulation tool.

### 2016-12-07 publications/Journal/2-background.tex
First changes after review
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything
Review abstract, intro, background
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Rewrote first part of the paper to the new synchronization terminology

This section briefly introduces the two most prominent synchronization protocols: conservative and optimistic synchronization.
Both algorithms are supported by Dxex.

>>>>

This section briefly introduces the two most prominent synchronization protocols: conservative and optimistic synchronization.
Both algorithms are supported by dxex. 
These protocols synchronize between kernels, but do not specify the parallelization of events within a single kernel.
For example, \pSim is orthogonal to this as it decides upon the parallelization of events within a single kernel.

### 2016-12-07 publications/Journal/3-features.tex
Rewrote first part of the paper to the new synchronization terminology

We will not make a detailed comparison with \textit{PythonPDEVS} here, but only mention some supported features.
\textit{Dxex} supports, similarly to \textit{PythonPDEVS}, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
We do not elaborate on these features in this paper.
But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints~\cite{PythonPDEVS_ACTIMS} to the simulation kernel.
In our case, a user can configure the simulation kernel with the synchronization protocol to use, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for (compiled) optimizations which were plainly impossible in an interpreted language.
\textit{Dxex} uses new optimizations from the C++14 standard when possible.
The C++11 standard threading primitives are used to run the different simulation kernels.
Within a single simulation kernel, OpenMP~\cite{openmp4} is used to parallelize the transition functions, as is usual in \textsf{Parallel DEVS}.

@@

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all domains, a general purpose simulation tool should support multiple protocols.
We argue that a general purpose simulation tool should support single core, conservative, and optimistic synchronization, as is the case for \textit{dxex}. Futhermore it should be possible to combine each inter kernel synchronization protocol with the parallel execution of events defined by \pSim.

These different protocols relate to three different model characteristics.
Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
Finally, single core simulation is still required for models where parallelism is bad. 
It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than single core.

Data exchange between different simulation kernels happens through shared memory, using the new C++11 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.

\subsubsection{Single Core}
Our single core simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
Minor modifications were made, though, such that it can be overloaded by different synchronization protocol implementations.
This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.
In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.

An overview of \textit{dxex}'s design is given in Figure~\ref{fig:class_diagram}.
It shows that there is a simulation \texttt{Kernel}, which simulates the \texttt{AtomicModel}s connected to it.
The superclass \texttt{Kernel} represents the single core simulation kernel.
Subclasses define specific variants, such as \texttt{ConservativeKernel} (conservative synchronization), \texttt{OptimisticKernel} (optimistic synchronization), and \texttt{DynamicKernel} (\textsf{Dynamic Structure DEVS}).

@@

\subsubsection{Conservative}
For conservative synchronization, each node must determine the nodes it is influenced by.
Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
Within the returned time interval, the model promises not to raise an event.

\subsubsection{Optimistic}
For optimistic synchronization, each node must be able to roll back to a previous point in time.
This is often implemented through the use of state saving.
This needs to be done carefully in order to avoid unnecessary copies, and minimize the overhead.
We use the default: explicitly save each and every intermediate state.
Mattern's algorithm~\cite{mattern} is used to determine the GVT, as it runs asynchronously and uses only $2n$ synchronization messages.
Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions (\textit{e.g.}, printing) are committed.

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local %(using \textit{tcmalloc})% see page 6 for introducing tcmalloc
 memory pools.
Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.

% protocol or inherent parallelism?
\subsubsection{\pSim}
In our tool each simulation kernel, regardless of inter kernel synchronization, is capable of executing concurrent transitions in parallel. A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions. The parallel execution of transitions introduces some overhead in thread pooling and locking, and disallows some optimizations in optimally rescheduling models. The concurrency of events and the combined computational load of the transitions has to outweigh this overhead to obtain a significant speedup, as we will demonstrate in Section~\textsc{\nameref{sec:4-performance}}.

\subsection{Synchronization Protocol Transparency}
We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be simulated using single core execution, conservative synchronization, or optimistic synchronization.
The synchronization protocol to use is a simple configuration option.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

Always defining a lookahead function might seem redundant, especially should the user never want to use conservative synchronization.
Defining the lookahead is difficult and can often be unpredictable.
The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
Depending on the model, simulation performance might still be faster than single core simulation.

Defining a lookahead function is therefore recommended in combination with conservative synchronization, but it is not a necessity, as a default value $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
Providing this default implementation has no impact in sequential or optimistic simulation; as the function is never called, the compiler will optimize it out.
By providing this default implementation in the model base class we ensure that a model can run in sequential, optimistic and conservative simulation by default.

\subsection{Increasing Parallelism}
The goal of our contribution is to increase simulation performance as much as possible, leveraging parallel computation in the process.
Parallelizing the simulation kernel goes further, however, than merely implementing the different synchronization protocols.

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for single core simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

Furthermore, allocating and deallocating memory by making calls to the operating system, as is typically done by calls such as \texttt{malloc}, happens sequentially.
To counter this, our memory allocators are backed by a thread-aware pooling library.
In a single core simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.

Intra-kernel events can be (de)allocated without requiring synchronization with the other kernels. 
They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation. 
\pSim therefore has an advantage in lower overhead in comparison with inter kernel synchronization protocols.
In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
As a result, simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators~\cite{Arena}.

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in multi core simulation kernels and arena-style allocators for single core simulation.
The latter can be faster, but at the cost of additional configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, reducing lock contention in \texttt{malloc} calls.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    \caption{Effect of memory allocators on multi core execution time, using optimistic synchronization.}
    \label{fig:memallocators_parallel}
\end{figure}

@@

We of course still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
\textit{Dxex} uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
Since the RNG is an implicit part of the state in the \textsf{Parallel DEVS} formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic \textsf{DEVS} model.

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator with optimistic synchronization.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s single core and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in single core simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
For multi core simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between single core and multi core simulation.
Conservative synchronization is almost unaffected by changing the RNG.

>>>>

We will not make a detailed comparison with \textit{PythonPDEVS} here, but only mention some supported features.
\textit{Dxex} supports, similarly to \textit{PythonPDEVS}, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
We do not elaborate on these features in this paper.
But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though not distributed).
This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints~\cite{PythonPDEVS_ACTIMS} to the simulation kernel.
In our case, a user can configure the simulation kernel with the synchronization protocol to use, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for (compiled) optimizations which were plainly impossible in an interpreted language, such as Python.
\textit{Dxex} uses new optimizations from the C++14 standard when possible.
The C++11 standard threading primitives are used to run the different simulation kernels.
Within a single simulation kernel, OpenMP~\cite{openmp4} is used to parallelize the transition functions, as is usual in \textsf{Parallel DEVS}.

@@

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all domains, a general purpose simulation tool should support multiple protocols.
We argue that a general purpose simulation tool should support all six standard synchronization protocols, as is the case for \textit{dxex}.

Different protocols relate to different model characteristics.
For example, \textit{Cx} is for when high lookahead exists between different nodes, whereas \textit{Ox} is for when lookahead is unpredictable.
It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than fully sequential execution (\textit{NN}).

Data exchange between different simulation kernels happens through shared memory, using the new C++11 synchronization primitives.
This was also possible in previous versions of the C++ standard by falling back to non-portable C functions.
C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.

\subsubsection{Inter-Kernel Synchronization}
Our core simulation algorithm is very similar to the one found in \textit{PythonPDEVS}, including many optimizations.
Minor modifications were made though, such that it can be overloaded by different synchronization protocol implementations.
This way, the \textsf{DEVS} simulation algorithm is implemented once, but parts can be overridden as needed.

An overview of \textit{dxex}'s design is given in Figure~\ref{fig:class_diagram}.
It shows that there is a simulation \texttt{Kernel}, which simulates the \texttt{AtomicModel}s connected to it.
The superclass \texttt{Kernel} represents a standalone simulation kernel (\textit{Nx}).
Subclasses define specific variants, such as \texttt{ConservativeKernel} (conservative synchronization), \texttt{OptimisticKernel} (optimistic synchronization), and \texttt{DynamicKernel} (\textsf{Dynamic Structure DEVS}).
In theory, more synchronization protocols (\textit{e.g.}, other algorithms for conservative synchronization) can be added without altering our design.

@@

The following inter-kernel synchronization protocols are implemented.

\paragraph{None (\textit{Nx})}
No inter-kernel synchronization is the base case, implemented in the \texttt{Kernel}.
It can be overloaded by any of the other simulation kernels, which augment it with inter-kernel synchronization methods.

\paragraph{Conservative (\textit{Cx})}
For conservative synchronization, each kernel determines the kernels it is influenced by.
Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
Within the returned time interval, the model promises not to raise an event.

\paragraph{Optimistic (\textit{Ox})}
For optimistic synchronization, each node must be able to roll back to a previous point in time.
This is implemented with state saving.
This needs to be done carefully in order to avoid unnecessary copies, and minimize the overhead.
We use the default: explicitly save each and every intermediate state.
Mattern's algorithm~\cite{mattern} is used to determine the GVT, as it runs asynchronously and uses only $2n$ synchronization messages.
Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions (\textit{e.g.}, printing) are committed.
The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local %(using \textit{tcmalloc})% see page 6 for introducing tcmalloc
 memory pools.
Again, we made use of specific new features of C++11, that are not available in Python, or even previous versions of the C++ language standard.

\subsubsection{Intra-Kernel Synchronization}
In our tool, each simulation kernel is capable of executing concurrent transitions in parallel, whether they are external, internal, or confluent.

\paragraph{None (\textit{xN})}
No intra-kernel parallelism is used, meaning that all concurrent transitions are processed sequentially.
Note that the order in which they are processed is non-deterministic.

\paragraph{\pSim (\textit{xP})}
With intra-kernel parallelism, a configurable number of threads is allocated to optimally divide the processing load of concurrent transitions.
The parallel execution of transitions introduces some overhead in thread pooling and locking, and disallows some optimizations in optimally rescheduling models.
The concurrency of events and the combined computational load of the transitions has to outweigh this overhead to obtain a significant speedup, as we will demonstrate in Section~\textsc{\nameref{sec:4-performance}}.

\subsection{Synchronization Protocol Transparency}
We define synchronization protocol transparency as having a single model, which can be executed on any kind of kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be simulated using any of the six synchronization protocols.
The synchronization protocol to use is a simple configuration option.
The exception is conservative synchronization (\textit{Cx}), where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

Always defining a lookahead function might seem redundant, especially should the user never want to use conservative synchronization.
Defining the lookahead is difficult and can often be unpredictable.
The more attractive option is for the simulation tool to provide a default lookahead function, such that simulation can run anyway, but likely not at peak performance.
Depending on the model, simulation performance might still be faster than \textit{Nx}.

Defining a lookahead function is therefore recommended in combination with conservative synchronization, but it is not a necessity, as a default value $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.
Providing this default implementation has no impact in sequential or optimistic simulation; as the function is never called, the compiler will optimize it out.
By providing this default implementation in the model base class we ensure that a model can run with any synchronization protocol.

\subsection{Increasing Parallelism}
The goal of our contribution is to increase simulation performance as much as possible, leveraging parallel computation in the process.
Parallelizing goes further, however, than merely implementing the different synchronization protocols.

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of CPU cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for \textit{Nx} synchronization will run correctly with conservative (\textit{Cx}) or optimistic (\textit{Ox}) synchronization without altering, from the point of view of the modeller, the (de)allocation semantics of events or states.

Furthermore, allocating and deallocating memory by making calls to the operating system, as is typically done by calls such as \texttt{malloc}, happens sequentially.
To counter this, our memory allocators are backed by a thread-aware pooling library.
With \textit{Nx} synchronization, no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.

Intra-kernel events can be (de)allocated without synchronization with the other kernels. 
They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation. 
In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur.
Intra-kernel synchronization protocols therefore have a lower overhead than inter-kernel synchronization protocols.
Simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators~\cite{Arena}.

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators for \textit{Cx} and \textit{Ox} synchronization, and arena-style allocators for \textit{Nx} synchronization.
The latter can be faster, but at the cost of additional configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, reducing lock contention in \texttt{malloc} calls.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    \caption{Effect of memory allocators with \textit{ON} synchronization.}
    \label{fig:memallocators_parallel}
\end{figure}

@@

We still need to guarantee determinism and isolation between the calls to the RNG, as well as avoiding excessive synchronization.
\textit{Dxex} uses the Tina RNG collection (TRNG)~\cite{PhysRevE.75.066701} as an alternative random number generator with performance and multithreading in mind.
%TODO one RNG per kernel or thread in new terminology?
Since the RNG is an implicit part of the state in the \textsf{Parallel DEVS} formalism, though often not implemented as such, we evaluated performance for both approaches: one global RNG per thread, and one RNG per atomic \textsf{DEVS} model.

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator with optimistic synchronization.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
Only \textit{Ox} synchronization seems affected, as it needs to copy more bytes in every transition due to state saving.
No additional copies need to happen in \textit{Nx} or \textit{Cx} synchronization.

Figure~\ref{fig:Queuerngspeedup} shows that, for \textit{NN} synchronization, any of the three alternatives is three times faster than STL RNG.
For \textit{Cx} and \textit{Ox} synchronization, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap with \textit{Nx} synchronization.
\textit{Cx} synchronization is almost unaffected by changing the RNG.
